# Workflow Basics

## Coding basics

No exercises.

## What’s in a name?

No exercises.

## Calling functions

No exercises.

## Practice

*1 - Why does this code not work?*

```{r eval = FALSE}
my_variable <- 10
my_varıable
#> Error in eval(expr, envir, enclos): object 'my_varıable' not found
```

*Look carefully! (This may seem like an exercise in pointlessness, but training your brain to notice even the tiniest difference will pay off when programming.)*

The value `10` is assigned to `my_variable`, not `my_varıable`. (Variables are named differently)

*2 - Tweak each of the following R commands so that they run correctly:*

```{r eval = FALSE}
library(tidyverse)

ggplot(dota = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))

fliter(mpg, cyl = 8)
filter(diamond, carat > 3)
```

Spelling is important. There are three typos in the above code:
- `dota` instead of `data`
- the first `filter` is misspelled as `fliter`
- the dataset `diamond` should be `diamonds`

The corrected code is shown below:

```{r eval = FALSE}
library(tidyverse)

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))

filter(mpg, cyl == 8)
filter(diamonds, carat > 3)
```

*3 - Press Alt + Shift + K. What happens? How can you get to the same place using the menus?*

The Keyboard Shortcut Quick Reference pops up. To get to the same place using the menus, click Help on the menu bar, and it is under Keyboard Shortcuts Help.
