#Vectors

## Introduction

No exercises.

## Vector basics

No exercises.

## Important types of atomic vector

### Exercises

```{r message = FALSE, warning = FALSE}
library(tidyverse)
```

*1 - Describe the difference between `is.finite(x)` and `!is.infinite(x)`.*

`is.finite(x)` returns `TRUE` if x is a finite value, hence not `NA`, `NaN`, or `Inf`.

```{r}
is.finite(c(123, NA, NaN, Inf))
```

`is.infinite(x)` returns `TRUE` if x is either `Inf` or `-Inf`. The `!` at the front means 'not', and changes the return value from `TRUE` to `FALSE`, or vice verse.

```{r}
!is.infinite(c(123, NA, NaN, Inf))
```

*2 - Read the source code for `dplyr::near()` (Hint: to see the source code, drop the ()). How does it work?*

```{r eval = FALSE}
function (x, y, tol = .Machine$double.eps^0.5) 
{
    abs(x - y) < tol
}
```

`.Machine$double.eps^0.5` is equal to 1.490116e-08. If the absolute difference between `x` and `y` is less than 1.490116e-08, then `x` and `y` are esentially equal, and the function returns `TRUE`.

*3 - A logical vector can take 3 possible values. How many possible values can an integer vector take? How many possible values can a double take? Use google to do some research.*

The largest possible integer is:

```{r}
.Machine$integer.max
```

Including negative integers, zero, and `NA`, the number of possible values an integer vector can take is:

```{r}
.Machine$integer.max * 2 + 1 + 1
```

```{r}
.Machine$double.xmax
```

*4 - Brainstorm at least four functions that allow you to convert a double to an integer. How do they differ? Be precise.*

`as.integer()` removes the decimal points and converts the type to integer:

```{r}
as.integer(11.23)
as.integer(-11.23)
typeof(as.integer(11.23))
```

`floor()` removes the decimal points and returns the largest integer not greater than the input values. However, the type of the value returned is double:

```{r}
floor(11.23)
floor(-11.23)
typeof(floor(11.23))
```

`trunc()` removes the decimal points and truncates the values toward 0. The returned value is double:

```{r}
trunc(11.23)
trunc(-11.23)
typeof(trunc(11.23))
```

There are also `ceiling()`, `round()`, and `signif()` that removes the decimal places and returns an integer. However, only `as.integer()` truly returns a value of type integer.

*5 - What functions from the readr package allow you to turn a string into logical, integer, and double vector?*

`parse_logical()`, `parse_integer()`, and `parse_double()`.

## Using atomic vectors

### Exercises

*1 - What does `mean(is.na(x))` tell you about a vector x? What about sum(!is.finite(x))?*

`mean(is.na(x))` tells you the proportion of values in a vector that are `NA`:

```{r}
x <- c(1,2,3,4,NA,NA, Inf)
mean(is.na(x))
```

`sum(!is.finite(x))` tells you the number of non-finite values in a vector:

```{r}
x <- c(1,2,3,4,NA,NA, Inf)
sum(!is.finite(x))
```

*2 - Carefully read the documentation of `is.vector()`. What does it actually test for? Why does `is.atomic()` not agree with the definition of atomic vectors above?*

`is.atomic()` does one simply job only, and that is the test whether the vector is an atomic vector:

```{r}
is.atomic(c(1,2,3,4))
is.atomic(c("Am", "I", "atomic?"))
```

`is.vector()` returns `TRUE` only if the vector has no attributes other than names:

```{r}
#vector with no attributes
x <- c(1,2,3,4)
#vector with names attribute
y <- c(x = 1, y = 2, z = 3, zz = 4)
#vector with two attributes: names and awesome
z <- c(x = 1, y = 2, z = 3, zz = 4)
attr(z, "awesome") <- "extreme"

is.vector(x)
is.vector(y)
is.vector(z)
```

*3 - Compare and contrast `setNames()` with `purrr::set_names()`*

In `setNames()`, you must provide the vector to be named and the vector of names.

In `purrr::set_names()`, three things can happen. You can provide the vector to be named and the vector of names, just like in `setNames()`.

```{r}
purrr::set_names(c(1,2,3), c("a", "b", "c"))
```

If the vector of names is not provided, the input vector is named after itself:

```{r}
purrr::set_names(c(1,2,3))
```

If the input vector is already named, a function can be provided to transform the original names:

```{r}
purrr::set_names(c(x = 1, y = 2, z = 3), toupper)
```

*4. Create functions that take a vector as input and returns:*

*The last value. Should you use `[` or `[[`?*

Using `[[` will ensure the last value will be returned even input vector is a list.

```{r}
fun1 <- function(x) {
  x[[length(x)]]
}

fun1(c(1,2,3,4,5))
```

*The elements at even numbered positions.*

```{r}
fun2 <- function(x) {
  x[seq_len(length(x)) %% 2 == 0]
}

fun2(c(1:10))
```

*Every element except the last value.*

```{r}
fun3 <- function(x){
  x[1:length(x) - 1]
}

fun3(1:10)
```

*Only even numbers (and no missing values).*

```{r}
fun4 <- function(x){
  iseven <- x %% 2 == 0
  nonNA <- !is.na(x)
  
  x[iseven & nonNA]
}

fun4(c(1:10, NA, NA, 11:16))
```

## Recursive vectors (lists)

### Exercise

*1. Draw the following lists as nested sets:*

*list(a, b, list(c, d), list(e, f))*

*list(list(list(list(list(list(a))))))*


*2. What happens if you subset a tibble as if you’re subsetting a list? What are the key differences between a list and a tibble?*

Subsetting a tibble with `[` will return a tibble:

```{r}
tblmtcars <- as.tibble(mtcars)

tblmtcars['cyl']
```

and `[[` will return a vector:

```{r}
tblmtcars[['cyl']]
```

Which is very similiar to subsetting in lists. (`[` returns a list, and `[[` returns a value inside the list)

## Attributes

No exercises.

## Augmented vectors

*1. What does hms::hms(3600) return? How does it print? What primitive type is the augmented vector built on top of? What attributes does it use?*

`hms::hms(3600)` returns the time in the format "hours:minutes:seconds".
```{r}
hms::hms(3600)
```

It is stored as a double, and the attributes are:

```{r}
attributes(hms::hms(3600))
```

*2. Try and make a tibble that has columns with different lengths. What happens?*

If the column has length of 1, it will be repeated:

```{r}
tibble(x = 1:6, y = 1)
```

Otherwise, if the columns are of different lengths, the shorter column will not be repeated and it will return an error:

```{r eval = FALSE}
> tibble(x = 1:6, y = 1:2)

# Error: Column `y` must be length 1 or 6, not 2
```

*3. Based on the definition above, is it ok to have a list as a column of a tibble?*

A list can have vectors of varying lengths, while a tibble must consist of columns of the same length. As long as the elements of the lists have the same length, otherwise it is not okay to have a list as a column of a tibble.