# Iteration

## Introduction

No exercises.

## For loops

### Exercises

```{r warning = FALSE, message = FALSE}
library(tidyverse)
```

*1. Write for loops to:*

*Compute the mean of every column in mtcars.*

```{r}
for (i in seq_along(mtcars)){
  print(paste("Mean of", colnames(mtcars)[i], ":", mean(mtcars[,i])))
}
```

*Determine the type of each column in nycflights13::flights.*

```{r}
for (i in seq_along(nycflights13::flights)){
  print(paste(colnames(nycflights13::flights)[i], ":",
              typeof(nycflights13::flights[[i]])))
}
```

*Compute the number of unique values in each column of iris.*

```{r}
for (i in seq_along(iris)){
  print(paste("Number of unique values in", colnames(iris)[i], ":",
              length(unique(iris[[i]]))))
}
```

*Generate 10 random normals for each of mean = -10, 0, 10, 100.*

```{r}
normal_means <- c(-10, 0, 10, 100)
for (i in seq_along(normal_means)){
  print(paste("mean:", normal_means[i]))
  print(rnorm(n = 10, mean = normal_means[i], sd = 1))
}
```

*2. Eliminate the for loop in each of the following examples by taking advantage of an existing function that works with vectors:*

```{r eval = FALSE}
out <- ""
for (x in letters) {
  out <- stringr::str_c(out, x)
}
```

is the same as:

```{r eval = FALSE}
stringr::str_c(letters, collapse = "")
```

```{r eval = FALSE}
x <- sample(100)
sd <- 0
for (i in seq_along(x)) {
  sd <- sd + (x[i] - mean(x)) ^ 2
}
sd <- sqrt(sd / (length(x) - 1))
```

is the same as:

```{r eval = FALSE}
x <- sample(100)
sd(x)
```

```{r eval = FALSE}
x <- runif(100)
out <- vector("numeric", length(x))
out[1] <- x[1]
for (i in 2:length(x)) {
  out[i] <- out[i - 1] + x[i]
}
```

is the same as:

```{r eval = FALSE}
cumsum(x)
```

*3. Combine your function writing and for loop skills:*

*Write a for loop that `prints()` the lyrics to the children's song "Alice the camel".*

```{r}
alice_the_camel <- function(x){
  for (i in x:0){
    if (i >= 2){
      writeLines(rep(paste("Alice the camel has", i, "humps."), 3))
    } else if (i == 1){
      writeLines(rep(paste("Alice the camel has", i, "hump."), 3))
    } else {
      writeLines(rep("Alice the camel has no humps.", 3))
    }
    
    if (i >= 1){
      writeLines("So go, Alice, go.\n")
    } else {
      writeLines("Now Alice is a horse")
    }
  }
}

alice_the_camel(5)
```

*Convert the nursery rhyme "ten in the bed" to a function. Generalise it to any number of people in any sleeping structure.*

```{r}
ten_in_the_bed <- function(x){
  for (i in x:1){
    writeLines(paste("There were", i, "in the bed"))
    writeLines("And the little one said,")
    if (i >= 2){
      writeLines("\"Roll over! Roll over!\"")
      writeLines("So they all rolled over and one fell out\n")
    } else {
      writeLines("\"Alone at last!\"")
    }
  }
}

ten_in_the_bed(5)
```

*Convert the song "99 bottles of beer on the wall" to a function. Generalise to any number of any vessel containing any liquid on any surface.*

```{r}
beer <- function(x){
  for (i in x:0){
    if (i >= 2){
      writeLines(paste(i, "bottles of beer on the wall," ,i, "bottles of beer."))
    } else if (i == 1){
      writeLines("1 bottle of beer on the wall, 1 bottle of beer.")
    } else {
      writeLines("No more bottles of beer on the wall, no more bottles of beer.")
    }
    
    if (i - 1 >= 2){
      writeLines(paste("Take one down and pass it around,", i - 1,
                       "bottles of beer on the wall.\n"))
    } else if (i - 1 == 1){
      writeLines("Take one down and pass it around, 1 bottle of beer on the wall.\n")
    } else if (i - 1 == 0){
      writeLines("Take one down and pass it around, no more bottles of beer on the wall.\n")
    } else{
      writeLines(paste("Go to the store and buy some more,", x, 
                       "bottles of beer on the wall."))
    }
  }
}

beer(5)
```

*4. It's common to see for loops that don't preallocate the output and instead increase the length of a vector at each step:*

```{r eval = FALSE}
output <- vector("integer", 0)
for (i in seq_along(x)) {
  output <- c(output, lengths(x[[i]]))
}
output
```

*How does this affect performance? Design and execute an experiment.*

As pointed in the text in late section, this is not very efficient because in each iteration, R has to copy all the data from the previous iterations.

Instead, create an empty list, store the values in the list, and flatten the list to a vector with `unlist()` at the end.

```{r eval = FALSE}
output <- vector("list", length(x))
for (i in seq_along(x)) {
  output[[i]] <- lengths(x[[i]])
}
unlist(output)
```

To Do.

## For loop variations

### Exercises

*1. Imagine you have a directory full of CSV files that you want to read in. You have their paths in a vector, `files <- dir("data/", pattern = "\\.csv$", full.names = TRUE)`, and now want to read each one with `read_csv()`. Write the for loop that will load them into a single data frame.*

```{r eval = FALSE}
files <- dir("data/", pattern = "\\.csv$", full.names = TRUE)
csv_list <- vector("list", length(files))

for (i in seq_along(files)){
  csv_list[[i]] <- read.csv(files[i])
}

dplyr::bind_rows(csv_list)

```

*2. What happens if you use for `(nm in names(x))` and `x` has no names? What if only some of the elements are named? What if the names are not unique?*

If `x` has no names:

```{r}
x <- c(1:10)

for (nm in names(x)){
  print(nm)
}
```

Nothing is printed, since `names(x)` returns `NULL`, the for loop does not run.

If only some of the elements are named:

```{r}
x <- c("a" = 1, 2, 3, "d" = 4)

for (nm in names(x)){
  print(nm)
}
```

The unnamed elements are given an empty character, and the number of iterations is equal to the length of the vector.

If the vector has non-unique names:

```{r}
x <- setNames(1:4, c("a","b","b","c"))

for (nm in names(x)){
  print(nm)
}
```

Again, the number of iterations is equal to the length of the vector.

*3. Write a function that prints the mean of each numeric column in a data frame, along with its name. For example, show_mean(iris) would print:*

```{r eval = FALSE}
show_mean(iris)
#> Sepal.Length: 5.84
#> Sepal.Width:  3.06
#> Petal.Length: 3.76
#> Petal.Width:  1.20
```

*(Extra challenge: what function did I use to make sure that the numbers lined up nicely, even though the variable names had different lengths?)*

```{r}
show_mean <- function(df){
  #empty vectors to store variable names and means
  start_sentence <- vector('character', ncol(df))
  col_means <- vector('double', ncol(df))
  
  #get names and means
  for (i in seq_len(ncol(df))){
    start_sentence[i] <- paste0(names(df)[i], ":")
    col_means[i] <- round(mean(df[[i]]),2)
  }
  
  #get max length of names and pad the shorter names
  start_sentence <- stringr::str_pad(start_sentence, max(nchar(start_sentence)),
                                     side = "right", pad = " ")
  
  #writeLines
  writeLines(paste(start_sentence, col_means))
}

show_mean(mtcars)
```

*4. What does this code do? How does it work?*

```{r eval = FALSE}
trans <- list( 
  disp = function(x) x * 0.0163871,
  am = function(x) {
    factor(x, labels = c("auto", "manual"))
  }
)
for (var in names(trans)) {
  mtcars[[var]] <- trans[[var]](mtcars[[var]])
}
```

A list can be used to store functions. The list `trans` stores two functions: `disp` to multiply the numerical values by a constant, and `am` creates a factor of two levels.

The for loop runs through the names in the list `trans` (in this case, only `disp` and `am`), then call the appropriate function, converts the values in the column and saves it.

## For loops vs. functionals

### Exercises

*1. Read the documentation for apply(). In the 2d case, what two for loops does it generalise?*

By rows and by columns.

*2. Adapt col_summary() so that it only applies to numeric columns. You might want to start with an is_numeric() function that returns a logical vector that has a TRUE corresponding to each numeric column.*

```{r}
col_summary <- function(df, fun){
  df2 <- df[,sapply(df, is.numeric)]

  start_sentence <- vector('character', ncol(df2))
  col_means <- vector('double', ncol(df2))
  
  #get names and means
  for (i in seq_len(ncol(df2))){
    start_sentence[i] <- paste0(names(df2)[i], ":")
    col_means[i] <- round(fun(df2[[i]]),2)
  }
  
  #get max length of names and pad the shorter names
  start_sentence <- stringr::str_pad(start_sentence, max(nchar(start_sentence)),
                                     side = "right", pad = " ")
  
  #writeLines
  writeLines(paste(start_sentence, col_means))
}

col_summary(iris, mean)
```

## The map functions

### Exercises

*1. Write code that uses one of the map functions to:*

*Compute the mean of every column in mtcars.*

```{r}
map_dbl(mtcars, mean) %>% round(2)
```

*Determine the type of each column in nycflights13::flights.*

```{r}
map_chr(nycflights13::flights, typeof)
```

*Compute the number of unique values in each column of iris.*

```{r}
map_int(iris, dplyr::n_distinct)
```

*Generate 10 random normals for each of mean = -10, 0, 10, 100*

```{r}
r_means <- c(-10, 0, 10, 100)

purrr::map(r_means, rnorm, n = 10)
```

*2. How can you create a single vector that for each column in a data frame indicates whether or not it's a factor?*

```{r}
map_lgl(mtcars, is.factor)
```

*3. What happens when you use the map functions on vectors that aren't lists? What does `map(1:5, runif)` do? Why?*

```{r}
purrr::map(1:5, runif)
```

It applies the function on each element in the vector. `map(1:5, runif)` applies the function `runif()` on 1, 2, 3, 4, and 5 individual, and return the results in a list (of length of 5).

*4. What does `map(-2:2, rnorm, n = 5)` do? Why? What does `map_dbl(-2:2, rnorm, n = 5)` do? Why?*
```{r}
purrr::map(-2:2, rnorm, n = 5)
```

`map(-2:2, rnorm, n = 5)` applies the function `rnorm()`, with the argument `n = 5` on each of the element in -2:2. `rnorm()` treats -2:2 as the input for its first remaining argument by position, which is the mean of the normal distribution. The function returns a list.

```{r eval = FALSE}
> map_dbl(-2:2, rnorm, n = 5)

#Error: Result 1 is not a length 1 atomic vector
```

`map_dbl(-2:2, rnorm, n = 5)` will not work because `map_dbl` returns an atomic vector. Each element can only store value of length of 1. If we change `n = 5` to `n = 1`:

```{r}
map_dbl(-2:2, rnorm, n = 1)
```

Then it will work.

*5. Rewrite `map(x, function(df) lm(mpg ~ wt, data = df))` to eliminate the anonymous function.*

```{r}
purrr::map(list(mtcars), ~lm(mpg ~ wt, data = .))
```

## Dealing with failure

No exercises.

## Mapping over multiple arguments

No exercises.

## Walk

No exercises

## Other patterns of for loops

### Exercises

*1. Implement your own version of `every()` using a for loop. Compare it with `purrr::every()`. What does purrr's version do that your version doesn't?*

```{r}
my_every <- function(x, .p){
  check <- sapply(x, .p)
  if (mean(check) == 1){
    return(TRUE)
  } else {
    return(FALSE)
  }
}

my_every(mtcars, is.numeric)
```

*2. Create an enhanced `col_sum()` that applies a summary function to every numeric column in a data frame.*

```{r}
col_sum <- function(df){
  df <- keep(df, is.numeric)
  
  return(purrr::map(df, summary))
}

col_sum(iris)
```

*3. A possible base R equivalent of `col_sum()` is:*

```{r eval = FALSE}
col_sum3 <- function(df, f) {
  is_num <- sapply(df, is.numeric)
  df_num <- df[, is_num]

  sapply(df_num, f)
}
```

*But it has a number of bugs as illustrated with the following inputs:*

```{r eval = FALSE}
df <- tibble(
  x = 1:3, 
  y = 3:1,
  z = c("a", "b", "c")
)
# OK
col_sum3(df, mean)
# Has problems: don't always return numeric vector
col_sum3(df[1:2], mean)
col_sum3(df[1], mean)
col_sum3(df[0], mean)
```

*What causes the bugs?*

The function fails if the filtered data frame has zero columns. Here is a modified version of `col_num3()` that will prevent the problem:

```{r eval = FALSE}
col_sum3 <- function(df, f) {
  is_num <- sapply(df, is.numeric)
  if (sum(is_num) >= 1){
    df_num <- df[, is_num]
    sapply(df_num, f)
  } else{
    NULL
  }
}
```

