# Tibbles

## Introduction

No exercises.

## Creating tibbles

No exercises.

## Tibbles vs. data.frame

No exercises.

## Interacting with older code

No exercises.

## Exercises

*1 - How can you tell if an object is a `tibble`? (Hint: try printing mtcars, which is a regular data frame).*

There are many ways to tell if an object is a `tibble`. We can look at the class type:

```{r}
class(mtcars)
```

Which shows `mtcars` is a `data.frame`. Printing `mtcars` also reveals that it's not `tibble` because it does not print 'A tibble' in the first row, it does not only show the first 10 rows, and all the columns do not fit on screen. Also, the type of each column is not reported.

*2 - Compare and contrast the following operations on a `data.frame` and equivalent `tibble`. What is different? Why might the default data frame behaviours cause you frustration?*

Creating `tibble` in R is pretty much the same as creating a `data.frame` object.

```{r message = FALSE, warning = FALSE}
library(tidyverse)
df <- data.frame(abc = 1, xyz = "a")
tdf <- tibble(abc = 1, xyz = "a")
```

However, in `data.frame`, strings are coerced into factors.

```{r}
str(df)
```

While in `tibble`, strings are still strings.

```{r}
str(tdf)
```

`tibble` does not do partial matching. It's perfectly fine to do this in `data.frame`:

```{r}
df$x
```

but we have to type the complete variable name in `tibble`:

```{r}
tdf$xyz
```

In `data.frame`, subsetting only one column with `[` `]`will return a vector, instead of a `data.frame` with one column:

```{r}
df[,'xyz']
```

In `tibble`, the same operation will return a `tibble` with a single column:

```{r}
tdf[,'xyz']
```

*3 - If you have the name of a variable stored in an object, e.g. `var <- "mpg"`, how can you extract the reference variable from a tibble?*

We will not be able to use `$` to subset the columns. Instead we need to use `[``]`.

```{r}
tibble_mtcars <- as.tibble(mtcars)
var <- 'mpg'
tibble_mtcars[var]
```

*4 - Practice referring to non-syntactic names in the following data frame by:*

```{r}
annoying <- tibble(
  `1` = 1:10,
  `2` = `1` * 2 + rnorm(length(`1`))
)
```

*1. Extracting the variable called `1`.*

```{r}
annoying$`1`
```

*2. Plotting a scatterplot of `1` vs `2`.*

```{r}
annoying %>% ggplot() +
  geom_point(mapping = aes(x = `1`, y = `2`))
```

*3. Creating a new column called `3` which is `2` divided by `1`.*

```{r}
annoying$`3` <- annoying$`2` / annoying$`1`
annoying
```

*4. Renaming the columns to `one`, `two` and `three`.*

```{r}
annoying %>% rename(one = `1`, two = `2`, three = `3`)
```

*5 - What does `tibble::enframe()` do? When might you use it?*

From the documentation, `enframe()` converts named atomic vectors or lists to two-column data frames. For unnamed vectors, the natural sequence is used as name column. For example:

```{r}
x <- c(Joe = 24, May = 33, Jack = 55)
enframe(x, name = 'Name', value = 'Age')
```

*6 - What option controls how many additional column names are printed at the footer of a tibble?*

By default, information of all remaining columns are printed at the footer. To limit the number of additional column information, we can use the argument `n_extra`. For exampe:

```{r}
print(nycflights13::flights, n_extra = 2)
```

