---
always_allow_html: yes
---
# Strings

## Introduction

No exercises

## String basics

### Exercises

```{r message = FALSE, warning = FALSE}
library(tidyverse)
library(stringr)
```

*1 - In code that doesn't use stringr, you'll often see `paste()` and `paste0()`. What's the difference between the two functions? What stringr function are they equivalent to? How do the functions differ in their handling of `NA`?*

`paste0()` is equivalent to `paste(..., sep = "")`. Both functions are equivalent to stringr function `str_c()`.

In `paste()` and`paste0`, `NA` are automaically converted to strings "NA". in `str_c()`, `NA` remains `NA`.

```{r}
paste('a', 'b', NA, 'c')
```
```{r}
paste0('a', 'b', NA, 'c')
```

```{r}
str_c('a', 'b', NA, 'c')
```

```{r}
str_c(str_replace_na(c('a', 'b', NA, 'c')), collapse = '')
```

*2 - In your own words, describe the difference between the `sep` and `collapse` arguments to str_c().*

One uses `sep` if combining different strings:

```{r}
str_c('a', 'b', 'c', sep = ' ')
```

If the lengths of the input strings are greater than 1, the operation is done element wise:

```{r}
str_c(1:5, letters[1:5], LETTERS[1:5], sep = ' ')
```

On the other hand, one uses `collapse` when combining the individual strings in a vector into a single string:

```{r}
str_c(c('a', 'b', 'c'), collapse = '/')
```

If `sep` and `collapse` are used together:

```{r}
str_c(1:5, letters[1:5], LETTERS[1:5], sep = ' ', collapse = '/')
```

*3 - Use `str_length()` and `str_sub()` to extract the middle character from a string. What will you do if the string has an even number of characters?*

If the string has an even number of characeters, let's assume we want to return both characters in the middle position.

We can write a simple function to do this job:

```{r}
middle_character <- function(s){
  s_length <- str_length(s)
  if (s_length %% 2 == 1) {
    return(str_sub(s, start = ceiling(s_length /2),
           end = ceiling(s_length / 2)))
  } else{
    return(str_sub(s, start = s_length /2,
           end = s_length / 2 + 1))
  }
}
```

A string with odd number of characeters:
```{r}
middle_character('level')
```

A string with even number of characeters:
```{r}
middle_character('middle')
```

*4 - What does `str_wrap()` do? When might you want to use it?*

In my own words, `str_wrap()` returns a long string of text into a "paragraph" format. At certain length, the function adds `\n` to insert a new line. Indent and exdent character width can also be specified.

For instance:

```{r}
long_text <- "This is the website for R for Data Science. This book will teach you how to do data science with R: You'll learn how to get your data into R, get it into the most useful structure, transform it, visualise it and model it. In this book, you will find a practicum of skills for data science. Just as a chemist learns how to clean test tubes and stock a lab, you'll learn how to clean data and draw plots-and many other things besides. These are the skills that allow data science to happen, and here you will find the best practices for doing each of these things with R. You'll learn how to use the grammar of graphics, literate programming, and reproducible research to save time. You'll also learn how to manage cognitive resources to facilitate discoveries when wrangling, visualising, and exploring data"

long_text_formatted <- str_wrap(long_text, width = 60, indent = 3)

long_text_formatted
```

Now we can use `writeLines()` to properly display the text:

```{r}
writeLines(long_text_formatted)
```

*5 - What does `str_trim()` do? What's the opposite of `str_trim()`?*

`str_trim()` trims whitesapce from start and end of string.

```{r}
str_trim(' white space     ')
```

The opposite if `str_pad()` adds whitespace, or other characeters, at the start and end of string.

```{r}
str_pad('interesting', width = 20, side = 'both', pad = '_')
```

*6 - Write a function that turns (e.g.) a vector `c("a", "b", "c")` into the string `a`, `b`, and `c`. Think carefully about what it should do if given a vector of length 0, 1, or 2.*

Let's call the function `custom_c`:

```{r}
custom_c <- function(s){
  if (length(s) == 0){
    return(NULL)
  } else if (length(s) == 1){
    return(s)
  } else if (length(s) == 2){
    return(str_c(s, collapse = ' and '))
  } else{
    first_half <- str_c(s[1:length(s) - 1], collapse = ', ')
    return(str_c(first_half, s[length(s)], sep = ', and '))
  }
}
```

Testing:

```{r}
print(custom_c(c()))
```

```{r}
print(custom_c(c('a')))
```

```{r}
print(custom_c(c('a','b')))
```

```{r}
print(custom_c(c('a','b','c','d','e')))
```

### Matching patterns with regular expressions

#### Basic matches

*1. Explain why each of these strings don't match a `\`: `"\"`, `"\\"`, `"\\\"`.*

In `"\"`, the single backlash "escapes" the special behavior of the close `"`, and it's waiting for the real second `"` to close the string. We can use `writeLines()` to demonstrate:

```{r echo = FALSE}
writeLines('> writeLines("\\")\n+')
```

The `+` is there waiting for more inputs.

In `"\\"`, the first backlash "escapes" the second backlash, so the second backlash does not "escape" the `"`. However, recall that we use strings to express regular expressions:

```{r}
writeLines("\\")
```

When evaluated, there is only one backlash, which is a special, escape character. Not the literal backlash.

In `"\\\"`, the third backlash "escapes" the `"`, which makes the input incomplete:

```{r echo = FALSE}
writeLines('> writeLines("\\\\\\\")\n+')
```

In `"\\\\"`, when evaluated:

```{r}
writeLines("\\\\")
```

It returns two backlashes. The first escapes the second, so the regular expression is now going to match the single backlash in the string sequence.

*2. How would you match the sequence `"'\`?*

```{r}
str_view("\"'\\", "\"\'\\\\")
```

*3. What patterns will the regular expression `\..\..\..` match? How would you represent it as a string?*

By now we should know that backlash itself is a special character and needs to be escaped in a string. To express the regular expression in a string, it should be `\\..\\..\\..`. What it does is it matches a dot, then any character, a dot, then any character, a dot, then finally any character.

```{r}
str_view(c(".c.o.m", "com", ".o.m.e.l"), "\\..\\..\\..")
```

#### Anchors

*1. How would you match the literal string `"$^$"`?*

```{r}
str_view("$^$", "\\$\\^\\$")
```

*2. Given the corpus of common words in `stringr::words`, create regular expressions that find all words that:*

*Start with "y".*

```{r}
str_view(stringr::words, "^y.*", match = TRUE)
```

To break down the regular expression `"^y.*"`, `^y` specifies that I want to look for a word that starts with a `y`, followed by any character `.`. `*` matches 0 or more `.`. The whole expression will match a string starting with a letter, followed by 0 or more other characters.

*End with "x"*.

```{r}
str_view(stringr::words, ".*x$", match = TRUE)
```

Again, `.*` specifies that it can start with any number of any characters, but `x$` states that it has to end with `x`.

*Are exactly three letters long. (Don't cheat by using str_length()!)*

```{r}
str_view(stringr::words, "^...$", match = TRUE)
```

*Have seven letters or more.*

```{r}
str_view(stringr::words, "^.{7,}", match = TRUE)
```

#### Character classes and alternatives

*1. Create regular expressions to find all words that:*

*Start with a vowel.*

```{r}
str_view(stringr::words, "^[aeiou].*", match = TRUE)
```

*That only contain consonants. (Hint: thinking about matching "not"-vowels.)*

```{r}
str_view(stringr::words, "^[^aeiou].*", match = TRUE)
```

*End with `ed`, but not with `eed`.*

```{r}
str_view(stringr::words, ".*([^e]ed)$", match = TRUE)
```

*End with `ing` or `ise`.*

```{r}
str_view(stringr::words, ".*(ing|ise)$", match = TRUE)
```

*2. Empirically verify the rule "i before e except after c".*

*3. Is "q" always followed by a "u"?*

```{r}
str_view(stringr::words, "(qu)", match = TRUE)
```

At least it seemes like it for the 980 words included in `stringr::words`.

*4. Write a regular expression that matches a word if it's probably written in British English, not American English.*

British:analyse. American:analyze.

```{r}
str_view(c("analyse", "analyze"), ".*(yse)$")
```


*5. Create a regular expression that will match telephone numbers as commonly written in your country.*

Numbers in Hong Kong, SAR:

```{r}
str_view(c("38888888", "3888-8888", "3888 8888"),
         "^\\d{4,4}\\s?-?\\d{4,4}$", match = TRUE)
```

#### Repetition

*1. Describe the equivalents of `?`, `+`, `\*` in `{m,n}` form.*

`?` - `{0,1}`
`+` - `{1,}`
`*` - `{0,}`

*2. Describe in words what these regular expressions match: (read carefully to see if I'm using a regular expression or a string that defines a regular expression.)*

`^.*$`

Matches a string of any length that starts with, ends with, and contains any characters, except a new line.

```{r}
str_view(c('interesting', 'what is this', 'a new line \n second line'), "^.*$")
```

`"\\{.+\\}"`

Matches a string with at least one character, enclosed by `{}`.

```{r}
str_view(c("{ok}", "(ok)"), "\\{.+\\}")
```

`\d{4}-\d{2}-\d{2}`

Matches a string of digits in this exact format: XXXX-XX-XX.

```{r}
str_view(c("1234-56-78", "1234-56-789"), "\\d{4}-\\d{2}-\\d{2}")
```


`"\\\\{4}"`

Matches exactly 4 `\`

```{r}
str_view(c("\\\\\\\\", "\\\\\\"), "\\\\{4}")
```

*3. Create regular expressions to find all words that:*

*Start with three consonants.*

`^[^aeiou]{3}`

*Have three or more vowels in a row.*

`[aeiou]{3,}`

*Have two or more vowel-consonant pairs in a row.*

`([aeiou][^aeiou]){2,}`

```{r}
str_view(c('ap', 'apap', 'apapap'), "([aeiou][^aeiou]){2,}")
```

#### Grouping and backreferences

*1. Describe, in words, what these expressions will match:*

`(.)\1\1` looks for two repeating characters:

```{r}
str_view(c("a","aa","aaa"), "(.)\\1")
```

`(.)(.)\\2\\1` matches strings in this pattern: xyyx

```{r}
str_view(c("awwa", "byb"), "(.)(.)\\2\\1")
```

`(..)\1` matches strings in this pattern: xyxy

```{r}
str_view("coco", "(..)\\1")
```

`"(.).\\1.\\1"` matches strings in this pattern: x.x.x

```{r}
str_view(c("a.a.a", "com.com.com"), "(.).\\1.\\1")
```

`"(.)(.)(.).*\\3\\2\\1"` matches strings that start with the pattern xyz, end with zyx, and any number of characters in between.

```{r}
str_view(c("abccba", "abclevelcba"), "(.)(.)(.).*\\3\\2\\1")
```

*2. Construct regular expressions to match words that:*

*Start and end with the same character.*

`"^(.).*\\1$"`

```{r}
str_view(c("level", "apple"), "^(.).*\\1$")
```

*Contain a repeated pair of letters (e.g. "church" contains "ch" repeated twice.)*

`"^(..).*\\1$"`

```{r}
str_view(c("level", "church"), "^(..).*\\1$")
```

*Contain one letter repeated in at least three places (e.g. "eleven" contains three "e"s.)*

`".*(.).*\\1.*\\1"`

```{r}
str_view(c("eleven", "level", "papaya"), ".*(.).*\\1.*\\1.*")
```

## Tools

### Detect matches

No exercises

### Exercises

*1. For each of the following challenges, try solving it by using both a single regular expression, and a combination of multiple `str_detect()` calls.*

*Find all words that start or end with x.*

```{r}
start_x <- str_detect(stringr::words, "^x")
end_x <- str_detect(stringr::words, "x$")

stringr::words[start_x | end_x]
```

*Find all words that start with a vowel and end with a consonant.*

```{r}
start_vowel <- str_detect(stringr::words, "^[aeiou]")
end_consonant <- str_detect(stringr::words, "[^aeiou]$")

stringr::words[start_vowel & end_consonant]
```

*Are there any words that contain at least one of each different vowel?*

```{r}
contains_a <- str_detect(stringr::words, "a")
contains_e <- str_detect(stringr::words, "e")
contains_i <- str_detect(stringr::words, "i")
contains_o <- str_detect(stringr::words, "o")
contains_u <- str_detect(stringr::words, "u")

stringr::words[contains_a & contains_e & contains_i & contains_o & contains_u]
```

Not in `stringr::words`.

*2. What word has the highest number of vowels? What word has the highest proportion of vowels? (Hint: what is the denominator?)*

```{r}
word_tbl <- tibble(
  words = stringr::words
) %>% 
  mutate(n_vowel = str_count(words, "[aeiou]"),
         n_consonant = str_count(words, "[^aeiou]"),
         n_chars = str_length(words),
         prop_vowel = n_vowel/n_chars,
         prop_consonant = n_consonant/n_chars)

word_tbl
```

Highest number of vowels:

```{r}
word_tbl %>%
  arrange(desc(n_vowel))
```

Highest proportion of vowels:

```{r}
word_tbl %>%
  arrange(desc(prop_vowel))
```

### Extract matches

#### Exercises

*1. In the previous example, you might have noticed that the regular expression matched "flickered", which is not a colour. Modify the regex to fix the problem.*

In the original code, the regular expression is:

```{r}
colours <- c("red", "orange", "yellow", "green", "blue", "purple")
colour_match <- str_c(colours, collapse = "|")
colour_match
```

It matches all strings that have one of the colour words, even words like `flickered`.

We can add `\b` to include word boundary. The modified regular expression:

```{r}
colours <- c("red", "orange", "yellow", "green", "blue", "purple")
colour_match <- str_c(colours, collapse = "|")
colour_match <- str_c("\\b(", colour_match, ")\\b", sep ='')
colour_match
```

```{r}
more <- sentences[str_count(sentences, colour_match) > 1]
str_view_all(more, colour_match)
```

*2. From the Harvard sentences data, extract:*

*The first word from each sentence.*

```{r}
str_extract(sentences, "^[A-Za-z\\']+") %>%
  head(10)
```

*All words ending in ing.*

```{r}
end_ing <- str_subset(sentences, "[A-Za-z]+ing\\b")

str_extract(end_ing, "[A-Za-z]+ing\\b") %>% unlist()
```

*All plurals.*

This is a little tough, just thinking how to extract all the irregular plural nouns (e.g., men, teeth). To make things easier, I'm extract words that have at least 4 letters and end with a `s`.

```{r}
plurals <- str_subset(sentences, "[A-Za-z]{3,}s\\b")

str_extract_all(plurals, "[A-Za-z]{3,}s\\b") %>% unlist() %>% head(20)
```

Not so bad... but not so good either. Let's stop here.

### Grouped matches

#### Exercises

*1. Find all words that come after a "number" like "one", "two", "three" etc. Pull out both the number and the word.*

```{r}
number_word <- "\\b(one|two|three|four|five|six|seven|eight|nine|ten)\\b\\s([A-Za-z]+)"

sentences[str_detect(sentences, number_word)] %>%
  str_match(number_word)
```

*2. Find all contractions. Separate out the pieces before and after the apostrophe.*

```{r}
contractions <- "\\b([A-Za-z]+)'([A-Za-z]+)"

sentences[str_detect(sentences, contractions)] %>%
  str_match(contractions)
```

### Replacing matches

#### Exercises

*1. Replace all forward slashes in a string with backslashes.*

```{r}
s <- str_replace_all("C:/Users/Someone", "/", "\\\\")
writeLines(s)
```

*2. Implement a simple version of str_to_lower() using replace_all().*

```{r}
str_replace_all("ABC", c("A" = "a",
                         "B" = "b",
                         "C" = "c"))
```

*3. Switch the first and last letters in `words`. Which of those strings are still words?*

```{r}
reversed_words <- words %>%
  str_replace_all("(.)(.+)(.)", "\\3\\2\\1")

words[reversed_words == words]
```

### Splitting

#### Exercises

*1. Split up a string like `"apples, pears, and bananas"` into individual components.*

```{r}
str_split("apples, pears, and bananas", "(, )|( )") %>%
  .[[1]]
```

*2. Why is it better to split up by `boundary("word")` than `" "`?*

Because there can be more than one empty spaces between the words.

```{r}
str_split("This is an       example.", " ") %>% .[[1]]
```

Using boundary, it will not be a problem.

```{r}
str_split("This is an       example.", boundary("word")) %>% .[[1]]
```

*3. What does splitting with an empty string ("") do? Experiment, and then read the documentation.*

```{r}
str_split("This is an example.", "") %>% .[[1]]
```

It splits the strings into individual characters.

### Find matches

No exercises

## Other types of pattern

### Exercises

*1. How would you find all strings containing \ with regex() vs. with fixed()?*

```{r}
str_view_all("\\", regex("\\\\"))
```

```{r}
str_view_all("\\", fixed("\\"))
```

*2. What are the five most common words in sentences?*

Top 10 most frequent words:

```{r}
sentences %>%
  str_split(boundary("word")) %>%
  unlist() %>%
  table() %>%
  sort(decreasing = TRUE) %>%
  head(10)
```

Top 10 most frequnet words

## Other uses of regular expressions

No exercises

## stringi

### Exercises

*1. Find the stringi functions that:*

*Count the number of words.*

`stringi::stri_count_words()`

```{r}
stringi::stri_count_words("This is awesome and \n great!.")
```

*Find duplicated strings.*

`stringi::stri_duplicated()` returns a vector of booleans, indicating which strings are duplicates.

```{r}
dup <- "this is super super super exciting."
dup <- stringr::str_split(dup, " ")[[1]]

stringi::stri_duplicated(dup)
```

Subset the duplicated strings:

```{r}
dup[stringi::stri_duplicated(dup)]
```

*Generate random text.*

`stringi::stri_rand_lipsum()`

```{r}
stringi::stri_rand_lipsum(1)
```

*2. How do you control the language that stri_sort() uses for sorting?*

By including the argument `locale = ....`