# Dates and times

## Introduction

No exercises.

## Creating date/times

### Exercises

```{r message = FALSE, warning = FALSE}
library(tidyverse)
library(lubridate)
library(nycflights13)
```

*1 - What happens if you parse a string that contains invalid dates?*

```{r}
ymd(c("2010-10-10", "bananas"))
```

It will return `NA`.

*2 - What does the `tzone` argument to `today()` do? Why is it important?*

`tzone` specifies which time zone you would like to find the current date of. It is important the dates of two different places in different timzones can differ.

*3 - Use the appropriate lubridate function to parse each of the following dates:*

```{r}
d1 <- "January 1, 2010"
mdy(d1)
```

```{r}
d2 <- "2015-Mar-07"
ymd(d2)

```

```{r}
d3 <- "06-Jun-2017"
dmy(d3)
```

```{r}
d4 <- c("August 19 (2015)", "July 1 (2015)")
mdy(d4)
```

```{r}
d5 <- "12/30/14" # Dec 30, 2014
mdy(d5)
```

## Date-time components

### Exercises

We use the `flights_dt` dataset (code is provided in the book) for the remaining exercises.

```{r}
make_datetime_100 <- function(year, month, day, time) {
  make_datetime(year, month, day, time %/% 100, time %% 100)
}

flights_dt <- flights %>% 
  filter(!is.na(dep_time), !is.na(arr_time)) %>% 
  mutate(
    dep_time = make_datetime_100(year, month, day, dep_time),
    arr_time = make_datetime_100(year, month, day, arr_time),
    sched_dep_time = make_datetime_100(year, month, day, sched_dep_time),
    sched_arr_time = make_datetime_100(year, month, day, sched_arr_time)
  ) %>% 
  select(origin, dest, ends_with("delay"), ends_with("time"))
```

*1 - How does the distribution of flight times within a day change over the course of the year?*

A distribution of flight times (departure by the hours) within a day looks something like this:

```{r}
flights_dt %>%
  mutate(date = make_date(year(dep_time),
                          month(dep_time),
                          mday(dep_time)),
         hour = hour(dep_time)) %>%
  group_by(date, hour) %>%
  filter(date == '2013-01-01') %>%
  ggplot(mapping = aes(x = hour)) +
  geom_density()
  
```

If we want to visualize the whole year:

```{r}
flights_dt %>%
  mutate(date = make_date(year(dep_time),
                          month(dep_time),
                          mday(dep_time)),
         hour = hour(dep_time)) %>%
  group_by(date, hour) %>%
  ggplot(mapping = aes(x = hour, group = date)) +
  geom_density(alpha = .1)
```

We see that the distributions for most days are very much the same.

*2 - Compare `dep_time`, `sched_dep_time` and `dep_delay`. Are they consistent? Explain your findings.*

Let's check if there are any inconsistent records:

```{r}
flights_dt %>% select(contains('dep')) %>%
  mutate(cal_delay = as.numeric(dep_time - sched_dep_time) / 60) %>%
  filter(dep_delay != cal_delay)
```

These inconsistent records have one thing in common. The `dep_delay` tells us the flights were delayed, but `cal_delay` tells us the flights departed early. The reason is that these delayed flights actually departed on the next day, and were not reflected in `dep_time`. If we add one day to the `dep_time`, the results should be consistent.

```{r}
flights_dt %>% select(contains('dep')) %>%
  mutate(cal_delay = as.numeric(dep_time - sched_dep_time) / 60) %>%
  filter(dep_delay != cal_delay) %>%
  mutate(dep_time = update(dep_time, mday = mday(dep_time) + 1)) %>%
  mutate(cal_delay = as.numeric(dep_time - sched_dep_time)) %>%
  filter(dep_delay != cal_delay)
```

All records are now consistent.

*3 - Compare `air_time` with the duration between the departure and arrival. Explain your findings. (Hint: consider the location of the airport.)*

To Do.

*4 - How does the average delay time change over the course of a day? Should you use `dep_time` or `sched_dep_time?` Why?*

We should use `sched_dep_time` because it will tell us how much delay we should expect at the scheduled departure time.

```{r}
flights_dt %>%
  mutate(hour = hour(sched_dep_time)) %>%
  group_by(hour) %>%
  summarize(avg_dep_delay = mean(dep_delay, na.rm = TRUE)) %>%
  ggplot(mapping = aes(x = hour, y = avg_dep_delay)) +
  geom_point() +
  geom_smooth(se = FALSE) +
  labs(y = "Average departure delay (in minutes)",
       x = "Scheduled departure time (in hours)")
```

*5 - On what day of the week should you leave if you want to minimise the chance of a delay?*

```{r}
flights_dt %>%
  mutate(dayweek = wday(sched_dep_time, label = TRUE)) %>%
  group_by(dayweek) %>%
  summarize(avg_dep_delay = mean(dep_delay, na.rm = TRUE),
            avg_arr_delay = mean(arr_delay, na.rm = TRUE)) %>%
  gather(key = 'delay', value = 'minutes', 2:3) %>%
  ggplot() +
  geom_col(mapping = aes(x = dayweek, y = minutes, fill = delay),
           position = 'dodge')
```

It looks like Saturdays are the best days to fly.

*6 - What makes the distribution of `diamonds$carat` and `flights$sched_dep_time` similar?*

In the distribution of `diamonds$carat`, we can see there are peaks at rounded numbers, like 0.3, 0.4, 0.5, 0.9, 1.0, 1.5, and 2.0.

```{r}
diamonds %>%
  ggplot() +
  geom_freqpoly(mapping = aes(x = carat), binwidth = .02)
```

In distribution of `flights_dt%sched_dep_time`, again we see peaks at rounded or nice numbers, such as 0, 5, 10, 15, etc.

```{r}
flights_dt %>%
  mutate(minutes = minute(sched_dep_time)) %>%
  ggplot() +
  geom_freqpoly(mapping = aes(x = minutes), binwidth = 1)
```

As the book suggests, these are most likely cuased by human factors, namely the convenience in representing nice, rounded numbers.

*7 - Confirm my hypothesis that the early departures of flights in minutes 20-30 and 50-60 are caused by scheduled flights that leave early. Hint: create a binary variable that tells you whether or not a flight was delayed.*

```{r}
flights_dt %>%
  mutate(delayed = dep_delay > 0,
         minutes = minute(sched_dep_time) %/% 10 * 10,
         minutes = factor(minutes, levels = c(0,10,20,30,40,50),
                          labels = c('0 - 9 mins',
                                     '10 - 19 mins',
                                     '20 - 29 mins',
                                     '30 - 39 mins',
                                     '40 - 49 mins',
                                     '50 - 50 mins'))) %>%
  group_by(minutes) %>%
  summarize(prop_early = 1 - mean(delayed, na.rm = TRUE)) %>%
  ggplot() +
  geom_point(mapping = aes(x = minutes, y = prop_early)) +
  labs(x = 'Scheduled departure (minutes)',
       y = 'Proportion of early departures')
```

## Time spans

### Exercises

*1 - Why is there `months()` but no `dmonths()`?*

Months do not have a fixed duration in seconds, unlike days, weeks, and years, because there are months with 28, 29, 30, and 31 days. 

*2 - Explain `days(overnight \* 1)` to someone who has just started learning R. How does it work?*

`overnight` is a boolean variable. If `arr_time` is less than `dep_time`, then the flight arrives on the next day, and `overnight` is `TRUE`; otherise, `FALSE`. Actually, the underlying value of `TRUE` is `1`, and `FALSE` is `0`, so the `* 0` can actually be omitted.

If `overnight` is `TRUE`, or `1`, `days(overnight)` will add one day to the `arr_time` and `sched_arr_time` datetime.

```{r}
flights_dt %>% 
  mutate(
    overnight = arr_time < dep_time,
    arr_time = arr_time + days(overnight),
    sched_arr_time = sched_arr_time + days(overnight)
  )
```

*3 - Create a vector of dates giving the first day of every month in 2015. Create a vector of dates giving the first day of every month in the current year.*

```{r}
make_date(year = 2015, month = 1:12, day = 1)
```

```{r}
make_date(year = year(today()), month = 1:12, day = 1)
```

*4 - Write a function that given your birthday (as a date), returns how old you are in years.*

```{r}
howold <- function(d) {
  age <- today() - d
  return(floor(age/dyears(1)))
}

howold(ymd(19860701))
```

*5 - Why can't `(today() %--% (today() + years(1))) / months(1)` work?*

It works?

```{r}
(today() %--% (today() + years(1))) / months(1)
```

The interval is defined as starting from today to a year from today, and there are 12 months in between.

## Time zones

No exercises.
